%% 中文摘要页
\begin{ChineseAbstract}
面向不确定环境中的机器人自主决策问题，安全强化学习在提升策略性能的同时，仍面临约束违背风险、感知误差影响以及训练环境与部署环境不一致等挑战；与此同时，小型多机器人系统具有部署灵活、适应性强等优势，但在多智能体协同中仍存在任务目标与安全约束相互耦合的难题。围绕安全性保障与多机器人协同两个核心问题，本文从安全控制、多机器人编队策略以及系统级验证平台三个层面展开研究，提出约束流形策略学习方法、几何约束策略学习方法以及容器化仿真验证平台，构成一套多机器人安全强化学习方法与工程化验证体系。

（1）针对单机器人自主决策中的安全保障问题，本文提出\textbf{约束流形策略学习方法}（CMPL, Constrained Manifold Policy Learning），由约束流形投影、安全可达性分析与数据驱动状态估计三部分构成统一框架。约束流形投影通过零空间映射将策略动作约束至安全流形的切空间，并结合奖励校准机制引导策略学习内生的安全行为；安全可达性分析基于历史轨迹数据预计算与环境相关的安全区域，在部署阶段无需在线重计算可达区域即可自适应调整安全裕度；数据驱动状态估计借助扩展卡尔曼滤波框架，由卷积神经网络学习观测噪声参数，使测量协方差能够随运动状态自适应调整，从而保证安全过滤器输入的可靠性。
实验结果表明，在目标到达任务消融实验中，CMPL 将平均安全代价由近端策略优化（PPO）基线方法的 64.59 降至 12.51，降幅达 80.6\%；在 Webots 仿真器中 E-puck 移动机器人零样本部署实验中，碰撞率由无安全过滤时的 40.0\% 降至在实验统计范围内未发生碰撞，位置误差由 0.051\,m 降至 0.037\,m，降幅 27.5\%。上述结果说明，CMPL 能够在显著降低约束违背风险的同时保持较好的迁移性与鲁棒性。

（2）针对多机器人编队导航中任务目标与安全约束相互耦合的情况，本文提出\textbf{几何约束策略学习方法}（GCPL, Geo-Constrained Policy Learning）。该方法以多智能体近端策略优化（MAPPO）为决策基础，在策略输出端引入几何约束过滤机制，对编队结构保持、机器人间安全距离以及避障要求进行统一处理，使安全性约束不再仅依赖奖励函数进行间接引导，而是在动作执行前完成显式修正。
实验结果显示，在线形编队任务中，GCPL 相较于 MAPPO 基线将安全代价由 105.78 降至 18.53，降幅 82.5\%；编队完成度由 10.00\% 提升至 71.00\%，成功率由 2.00\% 提升至 66.00\%。在圆形编队等复杂拓扑任务中，GCPL 将安全代价由 145.22 降至 34.40，降幅 76.3\%，成功率由 0.50\% 提升至 28.00\%。上述结果表明，GCPL 能够在保证安全性的同时显著提升任务完成效率与编队稳定性。

（3）为检验上述两类方法在不同场景中的可用性，并降低算法训练环境与部署环境不一致带来的影响，本文设计并实现了\textbf{面向多机器人安全强化学习的容器化仿真验证平台}。该平台采用分层架构组织仿真环境、算法接口、评估模块与结果展示模块，通过容器化方式统一运行依赖，减少不同实验配置之间的环境差异；同时设置标准化评估流程，对不同算法在相同任务条件下进行公平比较。在此基础上，本文将 CMPL 与 GCPL 接入平台，并从安全约束满足与任务完成效果等维度进行综合实验分析，进一步验证了 CMPL 与 GCPL 在统一仿真环境下的端到端集成能力与综合性能。

\ChineseKeywords{约束流形；多智能体强化学习；多机器人编队；安全强化学习}
\end{ChineseAbstract}

%% 英文摘要页
\begin{EnglishAbstract}
    For robot autonomous decision-making in uncertain environments, safe reinforcement learning still faces the risk of constraint violation, perception errors, and inconsistencies between training and deployment environments; meanwhile, small multi-robot systems offer flexible deployment and strong adaptability but suffer from coupling between task objectives and safety constraints during cooperation. Focusing on safety assurance and multi-robot coordination, this thesis investigates three aspects—safety control, multi-robot formation policy, and system-level verification platform—and proposes a Constrained Manifold Policy Learning method, a Geo-Constrained Policy Learning method, and a containerized simulation verification platform, forming a safe reinforcement learning method suite and its engineering verification framework.

    (1) For the safety assurance problem in single-robot autonomous decision-making, this thesis proposes the \textbf{Constrained Manifold Policy Learning} method (CMPL), a unified framework consisting of constrained-manifold projection, safe reachability analysis, and data-driven state estimation. The constrained-manifold projection maps policy actions onto the tangent space of the safety manifold via null-space projection and incorporates a reward calibration mechanism to drive the policy toward safe behavior. Safe reachability analysis pre-computes environment-dependent safety regions from historical trajectories so that the safety margin can be adapted at deployment without online recomputation of reachable regions. The data-driven state estimation employs an extended Kalman filter whose observation-noise parameters are learned by a convolutional neural network, allowing the measurement covariance to adapt with the motion state and ensuring reliable input to the safety filter. Experimental results show that on the goal-reaching ablation, CMPL reduces the average safety cost from 64.59 (PPO baseline) to 12.51, a reduction of 80.6\%; in zero-shot deployment of the E-puck robot in the Webots simulator, the collision rate decreases from 40.0\% (without safety filtering) to no observed collisions in the evaluation trials, and the position error from 0.051\,m to 0.037\,m, a 27.5\% reduction. These results demonstrate that the proposed method substantially reduces constraint-violation risk while preserving transferability and robustness.
        
    (2) For multi-robot formation navigation where task objectives and safety constraints are coupled, this thesis proposes the \textbf{Geo-Constrained Policy Learning} method (GCPL). Building on Multi-Agent Proximal Policy Optimization (MAPPO) as the decision backbone, GCPL introduces a geometric constraint filtering mechanism on the policy output that uniformly handles formation maintenance, inter-robot safe distance, and obstacle avoidance, so that safety constraints are no longer only indirectly steered by reward but are explicitly enforced before action execution.
    Experimental results show that on the line formation task, GCPL reduces the safety cost from 105.78 (MAPPO baseline) to 18.53 (an 82.5\% reduction), increases formation completion from 10.00\% to 71.00\%, and improves the success rate from 2.00\% to 66.00\%. On more complex topologies such as circle formation, GCPL reduces the safety cost from 145.22 to 34.40 (76.3\%) and increases the success rate from 0.50\% to 28.00\%. These results indicate that the proposed method significantly improves task efficiency and formation stability while preserving safety.
    
    (3) To verify the applicability of the above two methods across diverse scenarios and mitigate inconsistencies between training and deployment environments, this thesis designs and implements a \textbf{containerized simulation verification platform for multi-robot safe reinforcement learning}. The platform adopts a layered architecture covering simulation environments, algorithm interfaces, evaluation modules, and result visualization, and uses containerization to unify runtime dependencies and minimize environment drift across experiment configurations. A standardized evaluation pipeline allows fair comparison of different algorithms under identical task conditions. On this platform, this thesis integrates the safety constraint mechanism with multi-robot cooperative policies and analyzes their behavior in terms of constraint satisfaction, task completion, and formation consistency, thereby verifying the overall effectiveness of the proposed methods at the system level.

    \EnglishKeywords{Constraint Manifold; Multi-agent Reinforcement Learning; Multi-robot Formation; Safe Reinforcement Learning}
\end{EnglishAbstract}
